---
title: "Action execution lifecycle"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Action execution lifecycle}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

## Overview

`actionTypesR` turns ontology action definitions into real executions. Each submission
moves through the lifecycle: **resolve → validate → execute → record**.

This vignette uses the aviation demo bundle from `ontologySpecR` and an in-memory
DuckDB database to show the full flow.

## Setup

```{r}
library(ontologySpecR)
library(actionTypesR)

bundle <- read_bundle(system.file("examples", "aviation-demo.json", package = "ontologySpecR"))
con <- DBI::dbConnect(duckdb::duckdb())

DBI::dbWriteTable(con, "airports", data.frame(
  airport_id = c("DUB", "JFK"),
  name = c("Dublin", "John F Kennedy"),
  country = c("Ireland", "USA"),
  status = c("active", "active"),
  latitude = c(53.4213, 40.6413),
  longitude = c(-6.2701, -73.7781),
  stringsAsFactors = FALSE
))
```

## Register handlers

The aviation bundle includes `UpdateAirportStatus`, which is implemented in R. We can
register a handler directly in the context.

```{r}
ctx <- action_context(bundle, con)

ctx <- register_handler(ctx, "UpdateAirportStatus", function(conn, action, params, target_ids) {
  DBI::dbExecute(conn,
    "UPDATE airports SET status = ? WHERE airport_id = ?",
    params = list(params$new_status, target_ids[[1]])
  )
})
```

We can also add a new action that creates flight routes.

```{r}
create_route_action <- action_type(
  id = "CreateFlightRoute",
  targets = c("Airport"),
  parameters = list(
    parameter_def(id = "route_id", type = "string", required = TRUE),
    parameter_def(id = "origin", type = "string", required = TRUE),
    parameter_def(id = "destination", type = "string", required = TRUE),
    parameter_def(id = "status", type = "string", required = TRUE)
  ),
  effects = list(effect_def(kind = "create", object_type_id = "FlightRoute", notes = "")),
  impl_kind = "r",
  impl_entrypoint = "create_flight_route",
  policy = list(),
  extensions = list()
)

bundle$actions <- c(bundle$actions, list(create_route_action))
ctx <- action_context(bundle, con)

DBI::dbExecute(con, "CREATE TABLE routes (route_id TEXT, origin TEXT, destination TEXT, status TEXT)")

ctx <- register_handler(ctx, "CreateFlightRoute", function(conn, action, params, target_ids) {
  DBI::dbExecute(conn,
    "INSERT INTO routes (route_id, origin, destination, status) VALUES (?, ?, ?, ?)",
    params = list(params$route_id, params$origin, params$destination, params$status)
  )
})
```

## Resolve → validate → execute

```{r}
resolve_action(ctx, "UpdateAirportStatus")
```

```{r}
validate_params(ctx, "UpdateAirportStatus", list(new_status = "maintenance"))
```

```{r}
result <- submit_action(ctx, "UpdateAirportStatus",
  params = list(new_status = "maintenance"),
  target_ids = "DUB"
)
result
```

## Dry runs

Dry runs validate inputs and return an execution plan without dispatching the
handler or recording a log entry.

```{r}
submit_action(ctx, "UpdateAirportStatus",
  params = list(new_status = "closed"),
  target_ids = "JFK",
  .dry_run = TRUE
)
```

## Logs

Every successful or failed execution is recorded in the `action_log` table.

```{r}
action_log(ctx)
```

## Validation errors

Unknown parameters are rejected before execution.

```{r}
try(submit_action(ctx, "UpdateAirportStatus",
  params = list(wrong_param = "oops"),
  target_ids = "DUB"
))
```

## Policy checks

Policy functions receive `(action, params, target_ids, ctx)`. A 3-argument form
is also accepted for backward compatibility — it is wrapped automatically.

```{r}
ctx <- register_policy(ctx, function(action, params, target_ids) {
  if (params$new_status == "closed") {
    rlang::abort("Policy denied: airports cannot be closed during demo")
  }
  TRUE
})

try(submit_action(ctx, "UpdateAirportStatus",
  params = list(new_status = "closed"),
  target_ids = "DUB"
))
```

## Approval gates

An approval gate sits between the policy check and dispatch. The gate function
returns `"approved"`, `"denied"`, or `"pending"`. When `"pending"` is returned
the action is logged with `status = "pending_approval"` and no handler is called.

```{r}
# Reset policy so we can demonstrate the gate in isolation
ctx$policy_fn <- NULL

ctx <- register_approval_gate(ctx, function(action, params, target_ids, ctx) {
  if (identical(params$new_status, "maintenance")) {
    return("pending")   # requires human approval
  }
  "approved"
})

result <- submit_action(ctx, "UpdateAirportStatus",
  params = list(new_status = "maintenance"),
  target_ids = "DUB"
)
result$status   # "pending_approval"

# An approved action dispatches normally
result2 <- submit_action(ctx, "UpdateAirportStatus",
  params = list(new_status = "active"),
  target_ids = "DUB"
)
result2$status  # "success"
```

## Resubmitting a logged action

`resubmit_action()` looks up any prior submission by `submission_id` and
re-executes it, generating a fresh `submission_id`. The original log row is
never modified.

```{r}
second <- resubmit_action(ctx, result2$submission_id)
second$status  # "success"
```

## Log summary

`action_summary()` aggregates the log by action type and status, with optional
`action_type_id` and `since` filters.

```{r}
action_summary(ctx)
```

```{r}
DBI::dbDisconnect(con, shutdown = TRUE)
```
